# Getting Started with R and RStudio 2


This week, we will continue to explore R and RStudio.

```{r load, echo = F, message=FALSE, warning=FALSE}
library(tidyverse)
icebio <- read.delim(here::here("data", "icebio.txt"), sep = "\t", header = T)
```

## Working with tables

### Tabulating data

We can use the `table` function to create basic tables that extract raw frequency information. The following command tells us how many instances there are of each level of the variable `date` in the `icebio`. 

***

<div class="warning" style='padding:0.1em; background-color:#51247a; color:#f2f2f2'>
<span>
<p style='margin-top:1em; text-align:center'>
<b>TIP</b></p>
<p style='margin-left:1em;'>
</p></span>
</div>

<div class="question">` 

In order to access specific columns of a data frame, you can first type the name of the data set followed by a `$` symbol and then the name of the column (or variable).

</div>`

***


```{r, message=FALSE, warning=FALSE}
table(icebio$date) 
```

Alternatively, you could, of course, index the column by using its position in the data set like this: `icebio[, 6]` - the result of `table(icebio[, 6])` and `table(icebio$date)` are the same! Also note that here we leave out indexes for rows to tell R that we want all rows.

When you want to cross-tabulate columns, it is often better to use the `ftable` function (`ftable` stands for *frequency table*). 

```{r, message=FALSE, warning=FALSE}
ftable(icebio$age, icebio$sex)
```

***

<div class="warning" style='padding:0.1em; background-color:#51247a; color:#f2f2f2'>
<span>
<p style='margin-top:1em; text-align:center'>
<b>EXERCISE TIME!</b></p>
<p style='margin-left:1em;'>
</p></span>
</div>

<div class="question">` 

1. Using the `table` function, how many women are in the data collected between 2002 and 2005?

<details>
  <summary>Answer</summary>
  ```{r}
  table(icebio$date, icebio$sex)
  ```
</details>

2. Using the `ftable` function, how many men are are from northern Ireland in the data collected between 1990 and 1994?

<details>
  <summary>Answer</summary>
  ```{r}
  ftable(icebio$date, icebio$zone, icebio$sex)
  ```
</details>

</div>`

***


### Saving data to your computer

To save tabular data on your computer, you can use the `write.table`  function. This function requires the data that you want to save as its first argument, the location where you want to save the data as the second argument and the type of delimiter as the third argument. 

```{r savedisc, eval = F, message=FALSE, warning=FALSE}
write.table(icebio, here::here("data", "icebio.txt"), sep = "\t") 
```

**A word about paths**

In the code chunk above, the sequence `here::here("data", "icebio.txt")` is a handy way to define a path. A path is simply the location where a file is stored on your computer or on the internet (which typically is a server - which is just a fancy term for a computer - somewhere on the globe). The `here` function from the`here` package allows to simply state in which folder a certain file is and what file you are talking about. 

In this case, we want to access the file `icebio` (which is a `txt` file and thus has the appendix `.txt`) in the `data` folder. R will always start looking in the folder in which your project is stored. If you want to access a file that is stored somewhere else on your computer, you can also define the full path to the folder in which the file is. In my case, this would be `D:/Uni/UQ/SLC/LADAL/SLCLADAL.github.io/data`. However, as the `data` folder in in the folder where my Rproj file is, I only need to specify that the file is in the `data` folder within the folder in which my Rproj file is located.

**A word about package naming**

Another thing that is notable in the sequence `here::here("data", "icebio.txt")` is that I specified that the `here` function is part of the `here` package. This is what I meant by writing `here::here` which simply means use the `here` function from `here` package (`package::function`). This may appear to be somewhat redundant but it happens quite frequently, that different packages have functions that have the same names. In such cases, R will simply choose the function from the package that was loaded last. To prevent R from using the wrong function, it makes sense to specify the package AND the function (as I did in the sequence `here::here`). I only use functions without specify the package if the function is part of base R.

### Loading data from your computer

 To load tabular data from within your project folder (if it is in a tab-separated txt-file) you can also use the `read.delim` function. The only difference to loading from the web is that you use a path instead of a URL. If the txt-file is in the folder called *data* in your project folder, you would load the data as shown below. 

```{r, eval = T, message=FALSE, warning=FALSE}
icebio <- read.delim(here::here("data", "icebio.txt"), sep = "\t", header = T)
```

However, you can always just use the full path (and you must do this is the data is not in your project folder).

***

<div class="warning" style='padding:0.1em; background-color:#51247a; color:#f2f2f2'>
<span>
<p style='margin-top:1em; text-align:center'>
<b>NOTE</b><br>You may have to change the path to the data!</p>
<p style='margin-left:1em;'>
</p></span>
</div>

***

```{r, echo=T, eval=T, message=F, warning=F}
icebio <- read.delim(here::here("data", "icebio.txt"), 
                      sep = "\t", header = T)
```

To if this has worked, we will use the `head` function to see first 6 rows of the data

```{r, message=F, warning=F}
head(icebio)
```

### Loading Excel data

To load Excel spreadsheets, you can use the `read_excel` function from the `readxl` package as shown below. However, it may be necessary to install and activate the `readxl` package first. 

```{r, echo=T, eval=T, message=F, warning=F}
icebio <- readxl::read_excel(here::here("data", "ICEdata.xlsx"))
```

We now briefly check column names to see if the loading of the data has worked.

```{r, message=F, warning=F}
colnames(icebio)
```

### Loading text data

There are many functions that we can use to load text data into R. For example, we can use the `readLines` function as shown below.


```{r, echo=T, eval=T, message=F, warning=F}
text <- readLines(here::here("data", "text2.txt"))
# inspect first text element
text[1]
```

To load many texts, we can use a loop to read all texts in a folder as shown below. In a first step, we define the paths of the texts and then, we use the `sapply` function to loop over the paths and read them into R.


```{r load_texts, eval=T, echo=T, message=F, warning=F}
# define paths
paths <- list.files(here::here("data/testcorpus"), full.names = T)
# load texts
texts <- sapply(paths, function(x){ readLines(x) })
# inspect first text element
texts[1]
```

A method achieving the same result which uses piping (more on what that is below) and tidyverse R code is shown below.

```{r load_texts_tidy, eval=T, echo=T, message=F, warning=F}
# define paths
texts <- list.files(here::here("data/testcorpus"), full.names = T, pattern = ".*txt") %>%
  purrr::map_chr(~ readr::read_file(.))
# inspect first text element
texts[1]
```

### Renaming, Piping, and Filtering 

To rename existing columns in a table, you can use the `rename` command which takes the table as the first argument, the new name as the second argument, the an equal sign (=), and finally, the old name es the third argument. For example, renaming a column *OldName* as *NewName* in a table called *MyTable* would look like this: `rename(MyTable, NewName = OldName)`.  

Piping is done using the `%>%` sequence and it can be translated as **and then**. In the example below, we create a new object (icebio_edit) from the existing object (icebio) *and then* we rename the columns in the new object. When we use piping, we do not need to name the data we are using as this is provided by the previous step.

```{r, message=FALSE, warning=FALSE}
icebio_edit <- icebio %>%
  dplyr::rename(Id = id,
         FileSpeakerId = file.speaker.id,
         File = colnames(icebio)[3],
         Speaker = colnames(icebio)[4])
# inspect data
icebio_edit[1:5, 1:6]
```

A very handy way to rename many columns simultaneously, you can use the `str_to_title` function which capitalizes first letter of a word. In the example below, we capitalize all first letters of the column names of our current data.

```{r, message=FALSE, warning=FALSE}
colnames(icebio_edit) <- stringr::str_to_title(colnames(icebio_edit))
# inspect data
icebio_edit[1:5, 1:6]
```

To remove rows based on values in columns you can use the `filter` function.

```{r, message=FALSE, warning=FALSE}
icebio_edit2 <- icebio_edit %>%
  dplyr::filter(Speaker != "?",
         Zone != is.na(Zone),
         Date == "2002-2005",
         Word.count > 5)
# inspect data
head(icebio_edit2)
```

To select specific columns you can use the `select` function.

```{r, message=FALSE, warning=FALSE}
icebio_selection <- icebio_edit2 %>%
  dplyr::select(File, Speaker, Word.count)
# inspect data
head(icebio_selection)
```

You can also use the `select` function to remove specific columns.

```{r, message=FALSE, warning=FALSE}
icebio_selection2 <- icebio_edit2 %>%
  dplyr::select(-Id, -File, -Speaker, -Date, -Zone, -Age)
# inspect data
head(icebio_selection2)
```

### Ordering data

To order data, for instance, in ascending order according to a specific column you can use the `arrange` function.

```{r, message=FALSE, warning=FALSE}
icebio_ordered_asc <- icebio_selection2 %>%
  dplyr::arrange(Word.count)
# inspect data
head(icebio_ordered_asc)
```

To order data in descending order you can also use the `arrange` function and simply add a - before the column according to which you want to order the data.

```{r, message=FALSE, warning=FALSE}
icebio_ordered_desc <- icebio_selection2 %>%
  dplyr::arrange(-Word.count)
# inspect data
head(icebio_ordered_desc)
```

The output shows that the female speaker in file S2A-005 with the speaker identity A has the highest word count with 2,355 words. 

***

<div class="warning" style='padding:0.1em; background-color:#51247a; color:#f2f2f2'>
<span>
<p style='margin-top:1em; text-align:center'>
<b>EXERCISE TIME!</b></p>
<p style='margin-left:1em;'>
</p></span>
</div>

<div class="question">` 

1. Using the data called `icebio`, create a new data set called `ICE_Ire_ordered` and arrange the data in descending order by the number of words that each speaker has uttered. Who is the speaker with the highest word count?

<details>
  <summary>Answer</summary>
  ```{r}
  ICE_Ire_ordered <- icebio %>%
  dplyr::arrange(-word.count)
  # inspect data
  head(ICE_Ire_ordered)
  ```
</details>


</div>`

***


### Creating and changing variables

New columns are created, and existing columns can be changed, by using the `mutate` function. The `mutate` function takes two arguments (if the data does not have to be specified): the first argument is the (new) name of column that you want to create and the second is what you want to store in that column. The = tells R that the new column will contain the result of the second argument.

In the example below, we create a new column called *Texttype*. 

This new column should contain 

  + the value *PrivateDialoge* if *Filespeakerid* contains the sequence *S1A*, 
  
  + the value *PublicDialogue* if *Filespeakerid* contains the sequence *S1B*, 
  
  + the value *UnscriptedMonologue* if *Filespeakerid* contains the sequence *S2A*, 
  
  + the value *ScriptedMonologue* if *Filespeakerid* contains the sequence *S2B*, 
  
  + the value of *Filespeakerid* if *Filespeakerid* neither contains *S1A*, *S1B*, *S2A*, nor *S2B*.

```{r, message=FALSE, warning=FALSE}
icebio_texttype <- icebio_selection2 %>%
  dplyr::mutate(Texttype = 
                  dplyr::case_when(stringr::str_detect(Filespeakerid ,"S1A") ~ "PrivateDialoge",
                                   stringr::str_detect(Filespeakerid ,"S1B") ~ "PublicDialogue",
                                   stringr::str_detect(Filespeakerid ,"S2A") ~ "UnscriptedMonologue",
                                   stringr::str_detect(Filespeakerid ,"S2B") ~ "ScriptedMonologue",
                                   TRUE ~ Filespeakerid))
# inspect data
head(icebio_texttype)
```

### If-statements

We should briefly talk about if-statements (or `case_when` in the present case). The `case_when` function is both very powerful and extremely helpful as it allows you to assign values based on a test. As such, `case_when`-statements can be read as:

*When/If X is the case, then do A and if X is not the case do B!* (When/If -> Then -> Else)

The nice thing about `ifelse` or `case_when`-statements is that they can be used in succession as we have done above. This can then be read as:

*If X is the case, then do A, if Y is the case, then do B, else do Z* 

***

<div class="warning" style='padding:0.1em; background-color:#51247a; color:#f2f2f2'>
<span>
<p style='margin-top:1em; text-align:center'>
<b>EXERCISE TIME!</b></p>
<p style='margin-left:1em;'>
</p></span>
</div>

<div class="question">` 

1.Using the data called `icebio`, create a new data set called `ICE_Ire_AgeGroup` in which you create a column called `AgeGroup` where all speakers who are younger than 42 have the value *young* and all speakers aged 42 and over *old*. 

**Tip**: use if-statements to assign the *old* and *young* values.

<details>
  <summary>Answer</summary>
  ```{r}
  ICE_Ire_AgeGroup <- icebio %>%
  dplyr::mutate(AgeGroup = dplyr::case_when(age == "42-49" ~ "old",
                                            age == "50+" ~ "old", 
                                            age == "0-18" ~ "young",
                                            age == "19-25" ~ "young",
                                            age == "26-33" ~ "young",
                                            age == "34-41" ~ "young", 
                                            TRUE ~age))
  # inspect data
  head(ICE_Ire_AgeGroup); table(ICE_Ire_AgeGroup$AgeGroup)
  ```
</details>


</div>`

***

### Summarizing data

Summarizing is really helpful and can be done using the `summarise` function.

```{r, message=FALSE, warning=FALSE}
icebio_summary1 <- icebio_texttype %>%
  dplyr::summarise(Words = sum(Word.count))
# inspect data
head(icebio_summary1)
```

To get summaries of sub-groups or by variable level, we can use the `group_by` function and then use the `summarise` function.

```{r, warning=F, message=F}
icebio_summary2 <- icebio_texttype %>%
  dplyr::group_by(Texttype, Sex) %>%
  dplyr::summarise(Speakers = n(),
            Words = sum(Word.count))
# inspect data
head(icebio_summary2)
```


***

<div class="warning" style='padding:0.1em; background-color:#51247a; color:#f2f2f2'>
<span>
<p style='margin-top:1em; text-align:center'>
<b>EXERCISE TIME!</b></p>
<p style='margin-left:1em;'>
</p></span>
</div>

<div class="question">` 

1. Use the `icebio` and determine the number of words uttered by female speakers from Northern Ireland above an age of 50.

<details>
  <summary>Answer</summary>
  ```{r}
  words_fni50 <- icebio %>%
  dplyr::select(zone, sex, age, word.count) %>%
  dplyr::group_by(zone, sex, age) %>%
  dplyr::summarize(Words = sum(word.count)) %>%
  dplyr::filter(sex == "female",
         age == "50+",
         zone == "northern ireland")
  # inspect data
  words_fni50
  ```
</details>

2. Load the file *exercisedata.txt* and determine the mean scores of groups A and B.

**Tip**: to extract the mean, combine the `summary` function with the `mean` function. 

<details>
  <summary>Answer</summary>
  ```{r}
  exercisedata <- read.delim("data/exercisedata.txt", sep = "\t", header = T) %>%
  dplyr::group_by(Group) %>%
  dplyr::summarize(Mean = mean(Score))
  # inspect data
  exercisedata
  ```
</details>

</div>`

***


### Gathering and spreading data

The `tidyr` package has two very useful functions for gathering and spreading data that can be sued to transform data to long and wide formats (you will see what this means below). The functions are called `gather` and `spread`.

We will use the data set called `icebio_summary2`, which we created above, to demonstrate how this works.

We will first check out the `spread`-function to create different columns for women and men that show how many of them are represented in the different text types. 

```{r, message=FALSE, warning=FALSE}
icebio_summary_wide <- icebio_summary2 %>%
  dplyr::select(-Words) %>%
  tidyr::spread(Sex, Speakers)
# inspect
icebio_summary_wide
```

The data is now in what is called a `wide`-format as values are distributed across columns.

To reformat this back to a `long`-format where each column represents exactly one variable, we use the `gather`-function:

```{r, message=FALSE, warning=FALSE}
icebio_summary_long <- icebio_summary_wide %>%
  tidyr::gather(Sex, Speakers, female:male)
# inspect
icebio_summary_long
```



## Ending R sessions

At the end of each session, you can extract information about the session itself (e.g. which R version you used and which versions of packages). This can help others (or even your future self) to reproduce the analysis that you have done.

### Extracting session information

You can extract the session information by running the `sessionInfo` function (without any arguments)

```{r}
sessionInfo()
```

## Going further

If you want to know more, would like to get some more practice, or would like to have another approach to R, please check out the workshops and resources on R provided by the [UQ library](https://web.library.uq.edu.au/library-services/training). In addition, there are various online resources available to learn R (you can check out a very recommendable introduction [here](https://uvastatlab.github.io/phdplus/intror.html)). 

Here are also some additional resources that you may find helpful:

* Grolemund. G., and Wickham, H., [*R 4 Data Science*](http://r4ds.had.co.nz/), 2017.
    + Highly recommended! (especially chapters 1, 2, 4, 6, and 8)
* Stat545 - Data wrangling, exploration, and analysis with R. University of British Columbia.  <http://stat545.com/>
* Swirlstats, a package that teaches you R and statistics within R: <https://swirlstats.com/>
* DataCamp's (free) *Intro to R* interactive tutorial: <https://www.datacamp.com/courses/free-introduction-to-r>
    + DataCamp's advanced R tutorials require a subscription.
*Twitter: 
    + Explore RStudio Tips https://twitter.com/rstudiotips 
    + Explore #rstats, #rstudioconf



[Back to top](#getting_started_with_r_and_rstudio_2)




