# Logistic Regression and Model Diagnostics


This week, we continue with regression analyses and expand on last week by using a regression to analyse a nominal dependent variable. 

**Preparation and session set up**

This tutorial is based on R. If you have not installed R or are new to it, you will find an introduction to and more information how to use R [here](https://slcladal.github.io/intror.html). For this tutorials, we need to install certain *packages* from an R *library* so that the scripts shown below are executed without errors. Before turning to the code below, please install the packages by running the code below this paragraph. If you have already installed the packages mentioned below, then you can skip ahead and ignore this section. To install the necessary packages, simply run the following code - it may take some time (between 1 and 5 minutes to install all of the libraries so you do not need to worry if it takes some time).

```{r prep0, echo=T, eval = F, warning=F, message=F}
# install
install.packages("Boruta")
install.packages("car")
install.packages("emmeans")
install.packages("effects")
install.packages("flextable")
install.packages("ggplot2")
install.packages("ggpubr")
install.packages("Hmisc")
install.packages("knitr")
install.packages("lme4")
install.packages("MASS")
install.packages("mclogit")
install.packages("MuMIn")
install.packages("nlme")
install.packages("ordinal")
install.packages("rms")
install.packages("robustbase")
install.packages("sjPlot")
install.packages("stringr")
install.packages("tibble")
install.packages("dplyr")
install.packages("vcd")
install.packages("vip")
# install klippy for copy-to-clipboard button in code chunks
install.packages("remotes")
remotes::install_github("rlesur/klippy")
```

Now that we have installed the packages, we activate them as shown below.

```{r prep1, echo=T, eval=T, warning=F, message=F}
# set options
options(stringsAsFactors = F)          # no automatic data transformation
options("scipen" = 100, "digits" = 12) # suppress math annotation
# load packages
library(dplyr)
library(Boruta)
library(car)
library(effects)
library(emmeans)
library(flextable)
library(ggfortify)
library(ggplot2)
library(ggpubr)
library(Hmisc)
library(knitr)
library(lme4)
library(MASS)
library(mclogit)
library(MuMIn)
library(nlme)
library(ordinal)
library(rms)
library(robustbase)
library(sjPlot)
library(stringr)
library(tibble)
library(vcd)
library(vip)
# activate klippy for copy-to-clipboard button
klippy::klippy()
```


Once you have installed R and RStudio and initiated the session by executing the code shown above, you are good to go.


## Multiple Binomial Logistic Regression

Logistic regression is a multivariate analysis technique that builds on and is very similar in terms of its implementation to linear regression but logistic regressions take dependent variables that represent nominal rather than numeric scaling [@harrell2015regression]. The difference requires that the linear regression must be modified in certain ways to avoid producing non-sensical outcomes. The most fundamental difference between logistic and linear regressions is that logistic regression work on the probabilities of an outcome (the likelihood), rather than the outcome itself. In addition, the likelihoods on which the logistic regression works must be logged (logarithmized) in order to avoid produce predictions that produce values greater than 1 (instance occurs) and 0 (instance does not occur). You can check this by logging the values from -10 to 10 using the `plogis` function as shown below.

```{r blm0, message=FALSE, warning=FALSE}
round(plogis(-10:10), 5)
```

If we visualize these logged values, we get an S-shaped curve which reflects the logistic function.

```{r blm0b, echo = F, message=FALSE, warning=FALSE}
plogis(-10:10) %>%
  as.data.frame() %>%
  dplyr::mutate(id = 1:n()) %>%
  dplyr::rename(values = 1) %>%
  ggplot(aes(y = values, x = id)) +
  geom_line(color = "gray50", size = 1) +
  geom_point(color = "red", size = 2)+
  labs(x = "", y = "") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```


To understand what this mean, we will use a very simple example. In this example, we want to see whether the height of men affect their likelihood of being in a relationship. The data we use represents a data set consisting of two variables: height and relationship.

```{r blm1, echo=F, message=FALSE, warning=FALSE}
# set seed
set.seed(12345)
# generates 20 values, with mean of 30 & s.d.=2
bodyheight=rnorm(20,175,20) 
# sorts these values in ascending order
bodyheight=sort(bodyheight) 
# assign 'survival' to these 20 individuals non-randomly
# most mortality occurs at smaller body size
relationship=c(0,0,0,0,0,1,0,1,0,0,1,1,0,1,1,1,0,1,1,1) 
# saves data frame with two columns: body size, survival
blrdata=as.data.frame(cbind(bodyheight,relationship)) 
# generate models
lm1 <- lm(relationship ~ bodyheight, data = blrdata)
blr1 <- glm(relationship ~ bodyheight, family = binomial, data = blrdata)
blrdata <- blrdata %>%
  dplyr::mutate(Pred_lm = predict(lm1, blrdata),
                Pred_blm = predict(blr1, blrdata)) %>%
  dplyr::mutate(Prob_lm = plogis(predict(lm1, blrdata)),
                Prob_blm = plogis(predict(blr1, blrdata)),
                Pred_blm2 = ifelse(predict(blr1, blrdata, type = "response") >= .5, 1, 0))
# plot 
p1 <- ggplot(blrdata, aes(x = bodyheight, y =  relationship)) +
  geom_point() +
  geom_abline(intercept = coef(lm1)[1], slope = coef(lm1)[2], color = "red", size = .5) +
  geom_point(aes(x = bodyheight, y = Pred_lm), color = "blue") +
  theme_bw(base_size = 8)+
  coord_cartesian(ylim = c(-0.2, 1.2), xlim = c(125, 225)) +
  scale_y_continuous(breaks=seq(0, 1, 1), labels = c("Single", "Relationship")) +
  guides(fill = FALSE) +
  labs(title = "Predictions and \nregression line of lm.", x = "Height", y = "")

p2 <- ggplot(blrdata, aes(x = bodyheight, y =  relationship)) +
  geom_point() +
  geom_abline(intercept = coef(blr1)[1], slope = coef(blr1)[2], color = "red", size = .5) +
  geom_point(aes(x = bodyheight, y = Pred_blm), color = "blue") +
  theme_bw(base_size = 8)+
  coord_cartesian(ylim = c(-0.2, 1.2), xlim = c(125, 225)) +
  scale_y_continuous(breaks=seq(0, 1, 1), labels = c("Single", "Relationship")) +
  guides(fill = FALSE) +
  labs(title = "Predictions and \nregression line of blm.", x = "Height", y = "")


p3 <- ggplot(blrdata, aes(x = bodyheight, y =  relationship)) +
  geom_point() +
  geom_point(aes(x = bodyheight, y = Prob_blm), color = "blue") +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), 
              se = FALSE, color = "red", size = .5) +
  geom_segment(aes(xend = bodyheight, yend = Prob_blm), color = "red", alpha = .2) +
  theme_bw(base_size = 8) +
  coord_cartesian(ylim = c(-0.2, 1.2), xlim = c(125, 225)) +
  scale_y_continuous(breaks=seq(0, 1, 1), labels = c("Single", "Relationship")) +
  guides(fill = FALSE) +
  labs(title = "Logged Probabilities and \nlogged regression line of blm.", x = "Height", y = "")


# show plot
ggpubr::ggarrange(p1, p2, p3, nrow = 1)
```

The left panel of the Figure above shows that a linear model would predict values for the relationship status, which represents a factor (0 = Single and 1 = In a Relationship), that are nonsensical because values above 1 or below 0 do not make sense. In contrast to a linear regression, which predicts actual values, such as the frequencies of prepositions in a certain text, a logistic regression predicts *probabilities* of events (for example, being in a relationship) rather than actual values. The center panel shows the predictions of a logistic regression and we see that a logistic regression also has an intercept and a (very steep) slope but that the regression line also predicts values that are above 1 and below 0. However, when we log the predicted values we these predicted values are transformed into probabilities with values between 0 and 1. And the logged regression line has a S-shape which reflects the logistic function. Furthermore, we can then find the optimal line (the line with the lowest residual deviance) by comparing the sum of residuals - just as we did for a simple linear model and that way, we find the regression line for a logistic regression. 

### Example 1: EH in Kiwi English

To exemplify how to implement a logistic regression in R [see @agresti1996introduction; @agresti2011categorical] for very good and thorough introductions to this topic], we will analyze the use of the discourse particle *eh* in New Zealand English and test which factors correlate with its occurrence. The data set represents speech units in a corpus that were coded for the speaker who uttered a given speech unit, the gender, ethnicity, and age of that speaker and whether or not the speech unit contained an *eh*. To begin with, we clean the current work space, set option, install and activate relevant packages, load customized functions, and load the example data set.

```{r blm3}
# load data
blrdata  <- base::readRDS(url("https://slcladal.github.io/data/bld.rda", "rb"))
```

```{r blm4, echo = F}
# inspect data
blrdata %>%
  as.data.frame() %>%
  head(15) %>%
  flextable() %>%
  flextable::set_table_properties(width = .75, layout = "autofit") %>%
  flextable::theme_zebra() %>%
  flextable::fontsize(size = 12) %>%
  flextable::fontsize(size = 12, part = "header") %>%
  flextable::align_text_col(align = "center") %>%
  flextable::set_caption(caption = "First 15 rows of the blrdata.")  %>%
  flextable::border_outer()
```

The summary of the data show that the data set contains 25,821 observations of five variables. The variable `ID` contains strings that represent a combination file and speaker of a speech unit. The second variable represents the gender, the third the age, and the fourth the ethnicity of speakers. The fifth variable represents whether or not a speech unit contained the discourse particle *eh*. 

Next, we factorize the variables in our data set. In other words, we specify that the strings represent variable levels and define new reference levels because as a default `R` will use the variable level which first occurs in alphabet ordering as the reference level for each variable, we redefine the variable levels for Age and Ethnicity.

```{r blm5}
blrdata <- blrdata %>%
  # factorize variables
  dplyr::mutate(Age = factor(Age),
                Gender = factor(Gender),
                Ethnicity = factor(Ethnicity),
                ID = factor(ID),
                EH = factor(EH)) %>%
  # relevel Age (Reference = Young) and Ethnicity (Reference= Pakeha))
  dplyr::mutate(Age = relevel(Age, "Young"),
                Ethnicity = relevel(Ethnicity, "Pakeha"))
```

After preparing the data, we will now plot the data to get an overview of potential relationships between variables.

```{r blm6, message=FALSE, warning=FALSE}
blrdata %>%
  dplyr::mutate(EH = ifelse(EH == "0", 0, 1)) %>%
  ggplot(aes(Age, EH, color = Gender)) +
  facet_wrap(~Ethnicity) +
  stat_summary(fun = mean, geom = "point") +
  stat_summary(fun.data = mean_cl_boot, geom = "errorbar", width = 0.2) +
  theme_set(theme_bw(base_size = 10)) +
  theme(legend.position = "top") +
  labs(x = "", y = "Observed Probabilty of eh") +
  scale_color_manual(values = c("gray20", "gray70"))
```

With respect to main effects, the Figure above indicates that men use *eh* more frequently than women, that young speakers use it more frequently compared with old speakers, and that speakers that are descendants of European settlers (Pakeha) use *eh* more frequently compared with Maori (the native inhabitants of New Zealand).

The plots in the lower panels do not indicate significant interactions between use of *eh* and the Age, Gender, and Ethnicity of speakers. In a next step, we will start building the logistic regression model.

### Model Building

As a first step, we need to define contrasts and use the `datadist` function to store aspects of our variables that can be accessed later when plotting and summarizing the model. Contrasts define what and how variable levels should be compared and therefore influences how the results of the regression analysis are presented. In this case, we use treatment contrasts which are in-built. Treatment contrasts mean that we assess the significance of levels of a predictor against a baseline which is the reference level of a predictor. @field2012discovering [414-427] and @gries2021statistics provide very good and accessible explanations of contrasts and how to manually define contrasts if you would like to know more. 

```{r blm7, message=F, warning=F}
# set contrasts
options(contrasts  =c("contr.treatment", "contr.poly"))
# create distance matrix
blrdata.dist <- datadist(blrdata)
# include distance matrix in options
options(datadist = "blrdata.dist")
```

Next, we generate a minimal model that predicts the use of *eh* solely based on the intercept.

```{r blm8}
# baseline glm model
m0.glm = glm(EH ~ 1, family = binomial, data = blrdata)
```

### Model fitting

We will now start with the model fitting procedure. In the present case, we will use a manual step-wise step-up procedure during which predictors are added to the model if they significantly improve the model fit. In addition, we will perform diagnostics as we fit the model at each step of the model fitting process rather than after the fitting.

We will test two things in particular: whether the data has incomplete information or complete separation and if the model suffers from (multi-)collinearity. 

Incomplete information or complete separation means that the data does not contain all combinations of the predictor or the dependent variable. This is important because if the data does not contain cases of all combinations, the model will assume that it has found a perfect predictor. In such cases the model overestimates the effect of that that predictor and the results of that model are no longer reliable. For example, if *eh* was only used by young speakers in the data, the model would jump on that fact and say *Ha! If there is an old speaker, that means that that speaker will never ever and under no circumstances say *eh* - I can therefore ignore all other factors!* 

Multicollinearity means that predictors correlate and have shared variance. This means that whichever predictor is included first will take all the variance that it can explain and the remaining part of the variable that is shared will not be attributed to the other predictor. This may lead to reporting that a factor is not significant because all of the variance it can explain is already accounted for. However, if the other predictor were included first, then the original predictor would be returned as insignificant. This means that- depending on the order in which predictors are added - the results of the regression can differ dramatically and the model is therefore not reliable. Multicollinearity is actually a very common problem and there are various ways to deal with it but it cannot be ignored (at least in regression analyses).

We will start by adding *Age* to the minimal adequate model.

```{r blm9}
# check incomplete information
ifelse(min(ftable(blrdata$Age, blrdata$EH)) == 0, "not possible", "possible")
# add age to the model
m1.glm = glm(EH ~ Age, family = binomial, data = blrdata)
# check multicollinearity (vifs should have values of 3 or lower for main effects)
ifelse(max(vif(m1.glm)) <= 3,  "vifs ok", "WARNING: high vifs!") # VIFs ok
# check if adding Age significantly improves model fit
anova(m1.glm, m0.glm, test = "Chi")
```

As the data does not contain incomplete information, the vif values are below 3, and adding *Age* has significantly improved the model fit (the p-value of the ANOVA is lower than .05). We therefore proceed with *Age* included.

We continue by adding *Gender*. We add a second ANOVA test to see if including Gender affects the significance of other predictors in the model. If this were the case - if adding Gender would cause Age to become insignificant - then we could change the ordering in which we include predictors into our model.

```{r blm10}
ifelse(min(ftable(blrdata$Gender, blrdata$EH)) == 0, "not possible", "possible")
m2.glm <- update(m1.glm, . ~ . +Gender)
ifelse(max(vif(m2.glm)) <= 3,  "vifs ok", "WARNING: high vifs!") # VIFs ok
anova(m2.glm, m1.glm, test = "Chi")
Anova(m2.glm, test = "LR")
```


Again, including *Gender* significantly improves model fit and the data does not contain incomplete information or complete separation. Also, including *Gender* does not affect the significance of *Age*. Now, we include *Ethnicity*.


```{r blm11}
ifelse(min(ftable(blrdata$Ethnicity, blrdata$EH)) == 0, "not possible", "possible")
m3.glm <- update(m2.glm, . ~ . +Ethnicity)
ifelse(max(vif(m3.glm)) <= 3,  "vifs ok", "WARNING: high vifs!") # VIFs ok
anova(m3.glm, m2.glm, test = "Chi")
```

Since adding *Ethnicity* does not significantly improve the model fit, we do not need to test if its inclusion affects the significance of other predictors. We continue without *Ethnicity* and include the interaction between *Age* and *Gender*.

```{r blm12}
ifelse(min(ftable(blrdata$Age, blrdata$Gender, blrdata$EH)) == 0, "not possible", "possible")
m4.glm <- update(m2.glm, . ~ . +Age*Gender)
ifelse(max(vif(m4.glm)) <= 3,  "vifs ok", "WARNING: high vifs!") # VIFs ok
anova(m4.glm, m2.glm, test = "Chi")
```

The interaction between *Age* and *Gender* is not significant which means that men and women do not behave differently with respect to their use of `EH` as they age. Also, the data does not contain incomplete information and the model does not suffer from multicollinearity - the predictors are not collinear. We can now include if there is a significant interaction between *Age* and *Ethnicity*.

```{r blm13}
ifelse(min(ftable(blrdata$Age, blrdata$Ethnicity, blrdata$EH)) == 0, "not possible", "possible")
m5.glm <- update(m2.glm, . ~ . +Age*Ethnicity)
ifelse(max(vif(m5.glm)) <= 3,  "vifs ok", "WARNING: high vifs!") # VIFs ok
anova(m5.glm, m2.glm, test = "Chi")
```

Again, no incomplete information or multicollinearity and no significant interaction. Now, we test if there exists a significant interaction between *Gender* and *Ethnicity*.

```{r blm14}
ifelse(min(ftable(blrdata$Gender, blrdata$Ethnicity, blrdata$EH)) == 0, "not possible", "possible")
m6.glm <- update(m2.glm, . ~ . +Gender*Ethnicity)
ifelse(max(vif(m6.glm)) <= 3,  "vifs ok", "WARNING: high vifs!") # VIFs ok
anova(m6.glm, m2.glm, test = "Chi")
```


As the interaction between *Gender* and *Ethnicity* is not significant, we continue without it. In a final step, we include the three-way interaction between *Age*, *Gender*, and *Ethnicity*.

```{r blm15}
ifelse(min(ftable(blrdata$Age, blrdata$Gender, blrdata$Ethnicity, blrdata$EH)) == 0, "not possible", "possible")
m7.glm <- update(m2.glm, . ~ . +Gender*Ethnicity)
ifelse(max(vif(m7.glm)) <= 3,  "vifs ok", "WARNING: high vifs!") # VIFs ok
anova(m7.glm, m2.glm, test = "Chi")

```

We have found our final minimal adequate model because the 3-way interaction is also insignificant. As we have now arrived at the final minimal adequate model (m2.glm), we generate a final minimal model using the lrm function.

```{r blm16}
m2.lrm <- lrm(EH ~ Age+Gender, data = blrdata, x = T, y = T, linear.predictors = T)
m2.lrm
```


```{r blm17}
anova(m2.lrm)
```

After fitting the model, we validate the model to avoid arriving at a final minimal model that is overfitted to the data at hand.

### Model Validation

To validate a model, you can apply the `validate` function and apply it to a saturated model. The output of the `validate` function shows how often predictors are retained if the sample is re-selected with the same size but with placing back drawn data points. The execution of the function requires some patience as it is rather computationally expensive and it is, therefore, commented out below.

```{r blm18}
# model validation (remove # to activate: output too long for website)
m7.lrm <- lrm(EH ~ (Age+Gender+Ethnicity)^3, data = blrdata, x = T, y = T, linear.predictors = T)
#validate(m7.lrm, bw = T, B = 200)
```

The `validate` function shows that retaining two predictors (Age and Gender) is the best option and thereby confirms our final minimal adequate model as the best minimal model. In addition, we check whether we need to include a penalty for data points because they have too strong of an impact of the model fit. To see whether a penalty is warranted, we apply the `pentrace` function to the final minimal adequate model.

```{r blm19}
pentrace(m2.lrm, seq(0, 0.8, by = 0.05)) # determine penalty
```

The values are so similar  that a penalty is unnecessary. In a next step, we rename the final models.

```{r blm20}
lr.glm <- m2.glm  # rename final minimal adequate glm model
lr.lrm <- m2.lrm  # rename final minimal adequate lrm model
```

Now, we calculate a Model Likelihood Ratio Test to check if the final model performs significantly better than the initial minimal base-line model. The result of this test is provided as a default if we call a summary of the lrm object.

```{r blm21}
modelChi <- lr.glm$null.deviance - lr.glm$deviance
chidf <- lr.glm$df.null - lr.glm$df.residual
chisq.prob <- 1 - pchisq(modelChi, chidf)
modelChi; chidf; chisq.prob
```

The code above provides three values: a $\chi$^2^, the degrees of freedom, and a p-value. The p-value is lower than .05 and the results of the Model Likelihood Ratio Test therefore confirm that the final minimal adequate model performs significantly better than the initial minimal base-line model. Another way to extract the model likelihood test statistics is to use an ANOVA to compare the final minimal adequate model to the minimal base-line model.

A handier way to get these statistics is by performing an ANOVA on the final minimal model which, if used this way, is identical to a Model Likelihood Ratio test.

```{r blm22}
anova(m0.glm, lr.glm, test = "Chi") # Model Likelihood Ratio Test
```

In a next step, we calculate pseudo-R^2^ values which represent the amount of residual variance that is explained by the final minimal adequate model. We cannot use the ordinary R^2^ because the model works on the logged probabilities rather than the values of the dependent variable.


```{r blm23}
# calculate pseudo R^2
# number of cases
ncases <- length(fitted(lr.glm))
R2.hl <- modelChi/lr.glm$null.deviance
R.cs <- 1 - exp ((lr.glm$deviance - lr.glm$null.deviance)/ncases)
R.n <- R.cs /( 1- ( exp (-(lr.glm$null.deviance/ ncases))))
# function for extracting pseudo-R^2
logisticPseudoR2s <- function(LogModel) {
  dev <- LogModel$deviance
	nullDev <- LogModel$null.deviance
	modelN <-  length(LogModel$fitted.values)
	R.l <-  1 -  dev / nullDev
	R.cs <- 1- exp ( -(nullDev - dev) / modelN)
	R.n <- R.cs / ( 1 - ( exp (-(nullDev / modelN))))
	cat("Pseudo R^2 for logistic regression\n")
	cat("Hosmer and Lemeshow R^2  ", round(R.l, 3), "\n")
	cat("Cox and Snell R^2        ", round(R.cs, 3), "\n")
	cat("Nagelkerke R^2           ", round(R.n, 3),    "\n") }
logisticPseudoR2s(lr.glm)

```

The low pseudo-R^2^ values show that our model has very low explanatory power. For instance, the value of Hosmer and Lemeshow R^2^ (0.026) "is the proportional reduction in the absolute value of the log-likelihood measure and as such it is a measure of how much the badness of fit improves as a result of the inclusion of the predictor variables" [@field2012discovering 317]. In essence, all the pseudo-R^2^ values are measures of how substantive the model is (how much better it is compared to a baseline model). Next, we extract the confidence intervals for the coefficients of the model.

```{r blm24, message=F, warning=F}
# extract the confidence intervals for the coefficients
confint(lr.glm)
```

Despite having low explanatory and predictive power, the age of speakers and their gender are significant as the confidence intervals of the coefficients do not overlap with 0.

### Effect Size

In a next step, we compute odds ratios and their confidence intervals. Odds Ratios represent a common measure of effect size and can be used to compare effect sizes across models. Odds ratios rang between 0 and infinity. Values of 1 indicate that there is no effect. The further away the values are from 1, the stronger the effect. If the values are lower than 1, then the variable level correlates negatively with the occurrence of the outcome (the probability decreases) while values above 1 indicate a positive correlation and show that the variable level causes an increase in the probability of the outcome (the occurrence of EH).

```{r blm25}
exp(lr.glm$coefficients) # odds ratios
exp(confint(lr.glm))     # confidence intervals of the odds ratios
```

The odds ratios confirm that older speakers use *eh* significantly less often compared with younger speakers and that women use *eh* less frequently than men as the confidence intervals of the odds rations do not overlap with 1. In a next step, we calculate the prediction accuracy of the model.

### Prediction Accuracy

In order to calculate the prediction accuracy of the model, we generate a variable called *Prediction* that contains the predictions of pour model and which we add to the data. Then, we use the `confusionMatrix` function from the `caret` package [@caret] to extract the prediction accuracy. 


```{r blm26}
# create variable with contains the prediction of the model
blrdata <- blrdata %>%
  dplyr::mutate(Prediction = predict(lr.glm, type = "response"),
                Prediction = ifelse(Prediction > .5, 1, 0),
                Prediction = factor(Prediction, levels = c("0", "1")),
                EH = factor(EH, levels = c("0", "1")))
# create a confusion matrix with compares observed against predicted values
caret::confusionMatrix(blrdata$Prediction, blrdata$EH)
```

We can see that out model has never predicted the use of *eh* which is common when dealing with rare phenomena. This is expected as the event s so rare that the probability of it not occurring substantively outweighs the probability of it occurring. As such, the prediction accuracy of our model is not significantly better compared to the prediction accuracy of the baseline model which is the no-information rate (NIR)) (p = 0.5029).

We can use the `plot_model` function from the `sjPlot` package [@sjPlot] to visualize the effects.

```{r blm27, warning=F, message=F}
# predicted probability
efp1 <- plot_model(lr.glm, type = "pred", terms = c("Age"), axis.lim = c(0, 1)) 
# predicted percentage
efp2 <- plot_model(lr.glm, type = "pred", terms = c("Gender"), axis.lim = c(0, 1)) 
grid.arrange(efp1, efp2, nrow = 1)
```

And we can also combine the visualization of the effects in a single plot as shown below.

```{r blm28, message=F, warning=F}
sjPlot::plot_model(lr.glm, type = "pred", terms = c("Age", "Gender"), axis.lim = c(0, 1)) +
  theme(legend.position = "top") +
  labs(x = "", y = "Predicted Probabilty of eh", title = "") +
  scale_color_manual(values = c("gray20", "gray70"))
```


### Model Diagnostics

We are now in a position to perform model diagnostics and test if the model violates distributional requirements. In a first step, we test for the existence of multicollinearity.

### Multicollinearity

Multicollinearity means that predictors in a model can be predicted by other predictors in the model (this means that they *share* variance with other predictors). If this is the case, the model results are unreliable because the presence of absence of one predictor has substantive effects on at least one other predictor. 

To check whether the final minimal model contains predictors that correlate with each other, we extract variance inflation factors (VIF). If a model contains predictors that have variance inflation factors (VIF) > 10 the model is unreliable [@myers1990classical]. @gries2021statistics shows that a VIF of 10 means that that predictor is explainable (predictable) from the other predictors in the model with an R^2^ of .9 (a VIF of 5 means that predictor is explainable (predictable) from the other predictors in the model with an R^2^ of .8).Indeed, predictors with VIF values greater than 4 are usually already problematic but, for large data sets, even VIFs greater than 2 can lead inflated standard errors ([Jaeger 2013](http://wiki.bcs.rochester.edu/HlpLab/LSA2013Regression?action=AttachFile&do=view&target=LSA13-Lecture6-CommonIssuesAndSolutions.pdf)). Also, VIFs of 2.5 can be problematic [@szmrecsanyi2006morphosyntactic 215] and [@zuur2010protocol] proposes that variables with VIFs exceeding 3 should be removed.


<div class="warning" style='padding:0.1em; background-color:#f2f2f2; color:#51247a'>
<span>
<p style='margin-top:1em; text-align:center'>
<b>NOTE</b><br><br>However, (multi-)collinearity is only an issue if one is interested in interpreting regression results!  If the interpretation is irrelevant because what is relevant is prediction(!), then it does not matter if the model contains collinear predictors! See @gries2021statistics or the excursion below for a more elaborate explanation.</p>
<p style='margin-left:1em;'>
</p></span>
</div>

<br>


***

<div class="warning" style='padding:0.1em; background-color:#f2f2f2; color:#51247a'>
<span>
<p style='margin-top:1em; text-align:center'>
<b>EXCURSION</b></p>
<p style='margin-left:1em;'>
</p></span>
</div>

<div class="question">` 

What is multicollinearity?<br>

<details>
  <summary>Answer</summary>
  
  <br>
  
  During the workshop on mixed-effects modeling, we talked about (multi-)collinearity and someone asked if collinearity reflected shared variance (what I thought) or predictability of variables (what the other person thought). Both answers are correct! We will see below why...
  
  ***
  
  <div class="warning" style='padding:0.1em; background-color:#f2f2f2; color:#51247a'>
  <span>
  <p style='margin-top:1em; text-align:center'>
  (Multi-)collinearity reflects the predictability of predictors based on the values of other predictors!</p></span></div>
  
  
  ***
  
  To test this, I generate a data set with 4 independent variables `a`, `b`, `c`, and `d` as well as two potential response variables `r1` (which is random) and `r2` (where the first 50 data points are the same as in `r1` but for the second 50 data points I have added a value of 50 to the data points 51 to 100 from `r1`). This means that the predictors `a` and `d` should both strongly correlate with `r2`. 
  
  ```{r prep2, warning=F, message=F}
  # load packages
  library(dplyr)
  library(rms)
  # create data set
  # responses
  # 100 random numbers
  r1 <- rnorm(100, 50, 10)
  # 50 smaller + 50 larger numbers
  r2 <- c(r1[1:50], r1[51:100] + 50)
  # predictors
  a <- c(rep("1", 50), rep ("0", 50))
  b <- rep(c(rep("1", 25), rep ("0", 25)), 2)
  c <- rep(c(rep("1", 10), rep("0", 10)), 5)
  d <- c(rep("1", 47), rep ("0", 3), rep ("0", 47), rep ("1", 3))
  # create data set
  df <- data.frame(r1, r2, a, b, c, d)
  ```
  
  
  ```{r data1, echo = F, message=FALSE, warning=FALSE}
  # inspect data
  df %>%
    head(10) %>%
    kable(caption = "First 10 rows of df data.") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                              full_width = F)
  ```
  
  
  Here are the visualizations of r1 and r2
  
  ```{r data2, echo = F, message=FALSE, warning=FALSE}
  plot(r1, pch = 20)
  plot(r2, pch = 20)
  ```
  
  **Fit first model**
  
  Now, I fit a first model. As the response is random, we do not expect any of the predictors to have a significant effect and we expect the R^2^ to be rather low.
  
  ```{r}
  m1 <- lm(r1 ~ a + b + c + d, data = df)
  # inspect model
  summary(m1)
  ```
  
  We now check for (multi-)collinearity using the `vif` function from the `rms` package [@rms]. Variables `a` and `d` should have high variance inflation factor values (vif-values) because they overlap very much!
  
  ```{r}
  # extract vifs
  rms::vif(m1)
  ```
  
  Variables `a` and `d` do indeed have high vif-values.
  
  **Fit second model**
  
  We now fit a second model to the response which has higher values for the latter part of the response. Both `a` and `d` strongly correlate with the response. **But** because `a` and `d` are collinear, `d` should not be reported as being significant by the model. The R^2^ of the model should be rather high (given the correlation between the response r2 and `a` and `d`).
  
  ```{r}
  m2 <- lm(r2 ~ a + b + c + d, data = df)
  # inspect model
  summary(m2)
  ```
  
  Again, we extract the vif-values.
  
  ```{r}
  # extract vifs
  rms::vif(m2)
  ```
  
  The vif-values are identical which shows that what matters is if the variables are predictable. To understand how we arrive at vif-values, we inspect the model matrix.
  
  ```{r}
  # inspect model matrix
  mm <- model.matrix(m2)
  ```
  
  ```{r mm2, echo = F, message=FALSE, warning=FALSE}
  # inspect data
  mm %>%
    head(15) %>%
    kable(caption = "First 15 rows of the model matrix.") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                              full_width = F)
  ```
  
  We now fit a linear model in which we predict `d` from the other predictors in the model matrix.
  
  ```{r}
  mt <- lm(mm[,5] ~ mm[,1:4])
  summary(mt)$r.squared
  ```
  The R^2^ shows that the values of `d` are explained to 78.4 percent by the values of the other predictors in the model.
  
  Now, we can write a function [taken from @gries2021statistics] that converts this R^2^ value 
  
  ```{r}
  R2.to.VIF <- function(some.modelmatrix.r2) {
  return(1/(1-some.modelmatrix.r2)) } 
  R2.to.VIF(0.784)
  ```
  
  The function outputs the vif-value of `d`. This shows that the vif-value of `d` represents its predictability from the other predictors in the model matrix which represents the amount of shared variance between `d` and the other predictors in the model.
  
</details>

</div>`

***

<br>

We now extract and check the VIFs of the model.

```{r blm29}
vif(lr.glm)
```

In addition, predictors with 1/VIF values $<$ .1 must be removed (data points with values above .2 are considered problematic) [@menard1995applied] and the mean value of VIFs should be $~$ 1 [@bowerman1990linear].

```{r blm30}
mean(vif(lr.glm))
```

### Outlier detection

In order to detect potential outliers, we will calculate diagnostic parameters and add these to our data set.

```{r blm31}
infl <- influence.measures(lr.glm) # calculate influence statistics
blrdata <- data.frame(blrdata, infl[[1]], infl[[2]]) # add influence statistics
```

In a next step, we use these diagnostic parameters to check if there are data points which should be removed as they unduly affect the model fit.  


### Sample Size

We now check whether the sample size is sufficient for our analysis [@green1991many].

* if you are interested in the overall model: 50 + 8k (k = number of predictors)

* if you are interested in individual predictors: 104 + k

* if you are interested in both: take the higher value!

```{r blm32}
# function to evaluate sample size
smplesz <- function(x) {
  ifelse((length(x$fitted) < (104 + ncol(summary(x)$coefficients)-1)) == TRUE,
    return(
      paste("Sample too small: please increase your sample by ",
      104 + ncol(summary(x)$coefficients)-1 - length(x$fitted),
      " data points", collapse = "")),
    return("Sample size sufficient")) }
# apply unction to model
smplesz(lr.glm)
```

According to rule of thumb provided in @green1991many, the sample size is sufficient for our analysis.

### Summarizing Results

As a final step, we summarize our findings in tabulated form.

```{r blm33, message=F, warning=F}
sjPlot::tab_model(lr.glm)
```

***

A more detailed summary table can be retrieved as follows:

```{r blm34, eval = T, warning=F, message=F}
# load function
source("https://slcladal.github.io/rscripts/blrsummary.r")
# calculate accuracy 
predict.acc <- caret::confusionMatrix(blrdata$Prediction, blrdata$EH)
predict.acc <- predict.acc[3]$overall[[1]]
# create summary table
blrsummarytb <- blrsummary(lr.glm, lr.lrm, predict.acc) 
```

```{r blm35, echo = F, warning=F, message=F}
# inspect data
blrsummarytb %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Statistics") %>%
  flextable() %>%
  flextable::set_table_properties(width = .75, layout = "autofit") %>%
  flextable::theme_zebra() %>%
  flextable::fontsize(size = 12) %>%
  flextable::fontsize(size = 12, part = "header") %>%
  flextable::align_text_col(align = "center") %>%
  flextable::set_caption(caption = "Results of the binomial logistic regression analysis.")  %>%
  flextable::border_outer()
```

**R^2^ (Hosmer & Lemeshow)**

Hosmer and Lemeshow's R^2^ "is the proportional reduction in the absolute value of the log-likelihood  measure and as such it is a measure of how much the badness of fit improves  as a result of the inclusion of the predictor variables. It can vary between 0 (indicating that the predictors are useless at predicting the outcome variable) and 1 (indicating that the model predicts the outcome variable perfectly)" [@field2012discovering 317].

**R^2^ (Cox & Snell)**

 "Cox and Snell's R^2^ (1989) is based on the deviance of the model (-2LL(new»)
 and the deviance of the baseline model (-2LL(baseline), and the sample size,
 n [...]. However, this statistic never reaches its theoretical maximum of 1.

**R2 (Nagelkerke)**

 Since R^2^ (Cox & Snell) never reaches its theoretical maximum of 1,
 Nagelkerke (1991) suggested Nagelkerke's R^2^ [@field2012discovering 317-318].

**Somers’ D~xy~**

 Somers’ D~xy~ is a rank correlation between predicted probabilities and observed
 responses ranges between 0 (randomness) and 1 (perfect prediction). Somers' D~xy~ should have a value higher than .5 for the model to be meaningful [@baayen2008analyzing 204].

**C**

 C is an index of concordance between the predicted probability and the
 observed response. When C takes the value 0.5, the predictions are random,
 when it is 1, prediction is perfect. A value above 0.8 indicates that the
 model may have some real predictive capacity [@baayen2008analyzing 204].

**Akaike information criteria (AIC)**

Akaike information criteria (AlC = -2LL + 2k) provide a value that reflects a ratio between the number of predictors in the model and the variance that is explained by these predictors. Changes in AIC can serve as a measure of whether the inclusion of a variable leads to a significant increase in the amount of variance that is explained by the model. "You can think of this as the price you pay for something: you get a better value of R^2^, but you pay a higher price, and was that higher price worth it?  These information criteria help you to decide. The BIC is the same as the AIC but adjusts the penalty included in the AlC (i.e., 2k) by the number of cases: BlC = -2LL + 2k x log(n) in which n is the number of cases in the model" [@field2012discovering 318].

We can use the `reports` package [@report] to summarize the analysis.


```{r blr_report, message=F, warning = F}
report::report(lr.glm)
```

We can use this output to write up a final report: 

We fitted a logistic model (estimated using ML) to predict the use of the utterance-final discourse particle *eh* with Age and Gender (formula: EH ~ Age + Gender). The model's explanatory power is weak (Tjur's R^2^ = 0.03). The model's intercept, corresponding to Age = Young and Gender = Men, is at -0.23 (95% CI [-0.28, -0.19], p < .001). Within this model:

* The effect of Age [Old] is statistically significant and negative (beta = -0.83, 95% CI [-0.90, -0.77], p < .001; Std. beta = -0.83, 95% CI [-0.90, -0.77])

* The effect of Gender [Women] is statistically significant and negative (beta = -0.42, 95% CI [-0.47, -0.37], p < .001; Std. beta = -0.42, 95% CI [-0.47, -0.37])

Standardized parameters were obtained by fitting the model on a standardized version of the dataset. 95% Confidence Intervals (CIs) and p-values were computed using 


```{r fin}
sessionInfo()
```



[Back to top](#introduction)



