# Mixed-Effects Regression


This week, we will learn about mixed-effects regressions. So far, we have only learned about fixed-effects regression models but there are also mixed-effects regressions  (which are fitted using the `lme4` package [@lme4] in this tutorial). 

Fixed-effects regression models are models that assume a non-hierarchical data structure, i.e. data where data points are not nested or grouped in higher order categories (e.g. students within classes). The first part of this tutorial focuses on fixed-effects regression models while the second part focuses on mixed-effects regression models.

In contrast, mixed-effects models assume a hierarchical data structure in which data points are grouped or nested in higher order categories (e.g. students within classes or items within subjects). Mixed-effects models are rapidly increasing in use in data analysis because they allow us to incorporate hierarchical or nested data structures. Mixed-effects models are, of course, an extension of fixed-effects regression models and also multivariate and come in different types (e.g., linear, logistic, ordinal, multinomial). 

Mixed-effects models have the following advantages over simpler statistical tests: 

* Mixed-effects models are multivariate, i.e. they test the effect of several predictors simultaneously while controlling for the effect of all other predictors. 

* Mixed models allow to statistically incorporate within-speaker variability and are thus fit to model hierarchical or nested data structures. This applies if several observations are produced by an individual speaker, for instance.

* Mixed-models provide a wealth of diagnostic statistics which enables us to control e.g. (multi-)collinearity, i.e. correlations between predictors, and to test whether conditions or requirements are violated (e.g. homogeneity of variance, etc.). 

Major disadvantages of mixed-effects regression modeling are that they are prone to producing high $\beta$-errors [see @johnson2009getting] and that they require rather large data sets. 

The following focuses on an extension of ordinary multiple linear regressions: a mixed-effects linear regression. 

**Preparation and session set up**

For this week's content, we need to install certain *packages* from an R *library* so that the scripts shown below are executed without errors. hence, before turning to the code below, please install the packages by running the code below this paragraph - it may take some time (between 1 and 5 minutes to install all of the libraries so you do not need to worry if it takes some time).

```{r prep0, echo=T, eval = F, warning=F, message=F}
# install
install.packages("Boruta")
install.packages("car")
install.packages("emmeans")
install.packages("effects")
install.packages("flextable")
install.packages("ggplot2")
install.packages("ggpubr")
install.packages("Hmisc")
install.packages("knitr")
install.packages("lme4")
install.packages("MASS")
install.packages("mclogit")
install.packages("MuMIn")
install.packages("nlme")
install.packages("ordinal")
install.packages("rms")
install.packages("robustbase")
install.packages("sjPlot")
install.packages("stringr")
install.packages("tibble")
install.packages("dplyr")
install.packages("vcd")
install.packages("vip")
# install klippy for copy-to-clipboard button in code chunks
install.packages("remotes")
remotes::install_github("rlesur/klippy")
```

Now that we have installed the packages, we activate them as shown below.

```{r prep1, echo=T, eval=T, warning=F, message=F}
# set options
options(stringsAsFactors = F)          # no automatic data transformation
options("scipen" = 100, "digits" = 12) # suppress math annotation
# load packages
library(Boruta)
library(car)
library(effects)
library(emmeans)
library(flextable)
library(ggfortify)
library(ggplot2)
library(ggpubr)
library(Hmisc)
library(knitr)
library(lme4)
library(MASS)
library(mclogit)
library(MuMIn)
library(nlme)
library(ordinal)
library(rms)
library(robustbase)
library(sjPlot)
library(stringr)
library(tibble)
library(vcd)
library(vip)
# activate klippy for copy-to-clipboard button
klippy::klippy()
```


## Introduction

It is very common that the data is nested in the sense that data points are not independent because they are, for instance, produced by the same speaker or are grouped by some other characteristic. In such cases, the data is considered hierarchical and statistical models should incorporate such structural features of the data they work upon. Fortunately, modeling hierarchical or nested data structures is very easy thanks to the `lme4` package [@lme4].


With respect to regression modeling, hierarchical structures are incorporated by what is called *random effects*. When models only have a fixed-effects structure, then they make use of only a single intercept and/or slope (as in the left panel in the figure below), while mixed effects models have intercepts for each level of a random effect. If the random effect structure represents speakers then this would mean that a mixed-model would have a separate intercept and/or slope for each speaker (in addition to the overall intercept that is shown as an orange line in the figure below). 

```{r lmm1, echo=F, eval = T, message=FALSE, warning=FALSE}
Height <- c(169, 171, 164, 160, 158, 173, 166, 161, 180, 187, 170, 177, 163, 161, 157)
Weight <- c(68, 67, 65, 66, 64, 80, 75, 70, 85, 92, 86, 87, 85, 82, 80) 
Group <- c("a", "a", "a", "a", "a", "b", "b", "b", "b", "b", "c", "c", "c", "c", "c")
# create data sets
tb <- data.frame(Height,Weight, Group)
m1 <- lm(Weight ~ Height + Group, data = tb)
m2 <- lmer(Weight ~ Height + (1|Group), data = tb)
tb <- tb %>%
  dplyr::mutate(PWeight = predict(m1, tb),
                PWeight_lme = predict(m2, tb))
# plot
p1 <- ggplot(tb, aes(Height, Weight)) +
  geom_point(size = 2)
p2 <- ggplot(tb, aes(Height, Weight)) +
  geom_abline(intercept = summary(m2)$coefficients[1], slope = summary(m2)$coefficients[2], 
              color="orange", size = .75) +
  geom_point(size = 2) +
  ggtitle("Fixed-effects model \n(with fixed intercept)")

p3 <- ggplot(tb, aes(Height, Weight)) +
  geom_point(size = 2, aes(shape = Group, color = Group)) +
  geom_abline(intercept = fixef(m2)[1], slope = fixef(m2)[2], 
              color="orange", size = .75) +
  geom_smooth(method = "lm", se = F, aes(x = Height, y = PWeight, color = Group), size = .5) +
  theme(legend.position = "none") +
  ggtitle("Mixed-effects model \n (with random intercepts)")

p4 <- ggplot(tb, aes(Height, Weight)) +
  geom_smooth(se = F, method = "lm", size = .5, aes(shape = Group, color = Group))  +
  geom_abline(intercept = fixef(m2)[1], slope = fixef(m2)[2], 
              color="orange", size = .75) +
  geom_point(size = 2, aes(shape = Group, color = Group)) + 
  theme(legend.position = "none") +
  ggtitle("Mixed-effects model \n(with random intercepts + \nrandom slops)")
# show plot
ggpubr::ggarrange(p2, p3, p4, ncol = 3)
```

The idea behind regression analysis is expressed formally in the equation below where$f_{(x)}$ is the y-value we want to predict, $\alpha$ is the intercept (the point where the regression line crosses the y-axis at x = 0), $\beta$ is the coefficient (the slope of the regression line), and x is the value of a predictor (e.g. 180cm - if we would like to predict the weight of a person based on their height). The $\epsilon$ is an error term that reflects the difference between the predicted value and the (actually) observed value ($\epsilon$ is thus a residual that is important as regressions assume that residuals are, e.g., normally distributed). 

\begin{equation}
f_{(x)} = \alpha + \beta x + \epsilon
\end{equation}

In other words, to estimate how much someone weights who is 180 cm tall, we would multiply the coefficient (slope of the line) with 180 ($x$) and add the value of the intercept (point where line crosses the y-axis  at x = 0). 

The equation below represents a formal representation of a mixed-effects regression with varying intercepts [see @winter2019statistics, 235].

\begin{equation}
f_{(x)} = \alpha_{i} + \beta x + \epsilon
\end{equation}

In this random intercept model, each level of a random variable has a different intercept. To predict the value of a data point, we would thus take the appropriate intercept value (the model intercept + the intercept of the random effect) and add the product of the predictor coefficient and the value of x. 

Finally, the equation below represents a formal representation of a mixed-effects regression with varying intercepts and varying slopes [see @winter2019statistics, 235].

\begin{equation}
f_{(x)} = \alpha_{i} + \beta_{i}x + \epsilon
\end{equation}

In this last model, each level of a random variable has a different intercept and a different slope. To predict the value of a data point, we would thus take the appropriate intercept value (the model intercept + the intercept of the random effect) and add the coefficient of that random effect level multiplied by the value of x. 

## Random Effects

*Random Effects* can be visualized using two parameters: the intercept (the point where the regression line crosses the y-axis at x = 0) and the slope (the acclivity of the regression line). In contrast to fixed-effects models, that have only 1 intercept and one slope (left panel in the figure above), mixed-effects models can therefore have various *random intercepts* (center panel) or various *random slopes*, or both, various *random intercepts* and various *random slopes* (right panel). 

What features do distinguish random and fixed effects? 

1) Random effects represent a higher level variable under which data points are grouped. This implies that random effects must be categorical (or nominal but *they a´cannot be continuous*!) [see @winter2019statistics, p. 236].

2) Random effects represent a sample of an infinite number of possible levels. For instance, speakers, trials, items, subjects, or words represent a potentially infinite pool of elements from which many different samples can be drawn. Thus, random effects represent a random sample sample. Fixed effects, on the other hand, typically do not represent a random sample but a fixed set of variable levels (e.g. Age groups, or parts-of-speech).

3) Random effects typically represent many different levels while fixed effects typically have only a few. @zuur2013beginner propose that a variable may be used as a fixed effect if it has less than 5 levels while it should be treated as a random effect if it has more than 10 levels. Variables with 5 to 10 levels can be used as both. However, this is a rule of thumb and ignores the theoretical reasons (random sample and nestedness) for considering something as a random effect and it also is at odds with the way that repeated measures are models (namely as mixed effects) although they typically only have very few levels.  

4) Fixed effects represent an effect that if we draw many samples, the effect would be consistent across samples [@winter2019statistics] while random effects should vary for each new sample that is drawn.

In the following, we will only focus on models with random intercepts because this is the more common method and because including both random intercepts and random slopes requires larger data sets (but have a better fit because intercepts are not forced to be parallel and the lines therefore have a better fit). You should, however, always think about what random effects structure is appropriate for your model - a very recommendable explanation of how to chose which random effects structure is best (and about what the determining factors for this decision are) is give in @winter2019statistics[241-244]. Also, consider the center and the right plots above to understand what is meant by *random intercepts* and *random slopes*.

After adding random intercepts, predictors (or fixed effects) are added to the model (just like with multiple regression). So mixed-effects are called mixed-effects because they contain both random and fixed effects.

In terms of general procedure, random effects are added first, and only after we have ascertained that including random effects is warranted, we test whether including fixed-effects is warranted [@field2012discovering]. 

## Example: Preposition Use across Time by Genre

To explore how to implement a mixed-effects model in R we revisit the preposition data that contains relative frequencies of prepositions in English texts written between 1150 and 1913. As a first step, and to prepare our analysis, we load necessary R packages, specify options, and load as well as provide an overview of the data.

```{r lmm3}
# load data
lmmdata  <- base::readRDS(url("https://slcladal.github.io/data/lmd.rda", "rb")) %>%
  # convert date into a numeric variable
  dplyr::mutate(Date = as.numeric(Date))
```

```{r lmm3b, echo = F}
# inspect data
lmmdata %>%
  as.data.frame() %>%
  head(15) %>%
  flextable() %>%
  flextable::set_table_properties(width = .75, layout = "autofit") %>%
  flextable::theme_zebra() %>%
  flextable::fontsize(size = 12) %>%
  flextable::fontsize(size = 12, part = "header") %>%
  flextable::align_text_col(align = "center") %>%
  flextable::set_caption(caption = "First 15 rows of the lmmdata.")  %>%
  flextable::border_outer()
```

The data set contains the date when the text was written (*Date*), the genre of the text (*Genre*), the name of the text (*Text*), the relative frequency of prepositions in the text (*Prepositions*), and the region in which the text was written (*Region*). We now plot the data to get a first impression of its structure.

```{r lmm4, message=FALSE, warning=FALSE}
p1 <- ggplot(lmmdata, aes(x = Date, y = Prepositions)) +
  geom_point() +
  geom_smooth(method = "lm", se = F, color = "red", linetype = "dashed") +
  theme_bw() +
  labs(y = "Frequency\n(Prepositions)")
p2 <- ggplot(lmmdata, aes(x = reorder(Genre, -Prepositions), y = Prepositions)) +
  geom_boxplot() +
  theme_bw() + 
  theme(axis.text.x = element_text(angle=90)) +
  labs(x = "Genre", y = "Frequency\n(Prepositions)")
p3 <- ggplot(lmmdata, aes(Prepositions)) +
  geom_histogram() +
  theme_bw() + 
  labs(y = "Count", x = "Frequency (Prepositions)")
grid.arrange(grobs = list(p1, p2, p3), widths = c(1, 1), layout_matrix = rbind(c(1, 1), c(2, 3)))
```


The scatter plot in the upper panel indicates that the use of prepositions has moderately increased over time while the boxplots in the lower left panel show that the genres differ quite substantially with respect to their median frequencies of prepositions per text. Finally, the histogram in the lower right panel show that preposition use is distributed normally with a mean of 132.2 prepositions per text. 

```{r lmm5, message=FALSE, warning=FALSE}
p4 <- ggplot(lmmdata, aes(Date, Prepositions)) +
  geom_point() +
  labs(x = "Year", y = "Prepositions per 1,000 words") +
  geom_smooth(method = "lm")  + 
  theme_bw()
p5 <- ggplot(lmmdata, aes(Region, Prepositions)) +
  geom_boxplot() +
  labs(x = "Region", y = "Prepositions per 1,000 words") +
  geom_smooth(method = "lm")  + 
  theme_bw()
grid.arrange(p4, p5, nrow = 1)
```

```{r lmm6, message=FALSE, warning=FALSE}
ggplot(lmmdata, aes(Date, Prepositions)) +
  geom_point() +
  facet_wrap(~ Genre, nrow = 4) +
  geom_smooth(method = "lm") +
  theme_bw() +
  labs(x = "Date of composition", y = "Prepositions per 1,000 words") +
  coord_cartesian(ylim = c(0, 220))
```

Centering or even scaling numeric variables is useful for later interpretation of regression models: if the date variable were not centered, the regression would show the effects of variables at year 0(!). If numeric variables are centered, other variables are variables are considered relative not to 0 but to the mean of that variable (in this case the mean of years in our data). Centering simply means that the mean of the numeric variable is subtracted from each value.

```{r lmm7}
lmmdata$DateUnscaled <- lmmdata$Date
lmmdata$Date <- scale(lmmdata$Date, scale = F)
```


```{r lmm7b, echo = F}
# inspect data
lmmdata %>%
  as.data.frame() %>%
  head(15) %>%
  flextable() %>%
  flextable::set_table_properties(width = .75, layout = "autofit") %>%
  flextable::theme_zebra() %>%
  flextable::fontsize(size = 12) %>%
  flextable::fontsize(size = 12, part = "header") %>%
  flextable::align_text_col(align = "center") %>%
  flextable::set_caption(caption = "First 15 rows of the lmmdata.")  %>%
  flextable::border_outer()
```

We now set up a fixed-effects model with the `glm` function and a mixed-effects model using the `glmer` function from the `lme4` package [@lme4] with Genre as a random effect.

```{r lmm8, message=FALSE, warning=FALSE}
# generate models
m0.glm <- glm(Prepositions ~ 1, family = gaussian, data = lmmdata)
m0.lmer = lmer(Prepositions ~ 1 + (1|Genre), REML = T, data = lmmdata)
```

Now that we have created the base-line models, we will test whether including a random effect structure is mathematically justified. It is important to note here that we are not going to test if including a random effect structure is theoretically motivated but simply if it causes a decrease in variance.


## Testing Random Effects

As a first step in the modeling process, we now need to determine whether or not including a random effect structure is justified. We do so by comparing the AIC of the base-line model without random intercepts to the AIC of the model with random intercepts. 


```{r lmm9a, message=F, warning=F}
AIC(logLik(m0.glm))
AIC(logLik(m0.lmer))
```

The inclusion of a random effect structure with random intercepts is justified as the AIC of the model with random intercepts is substantially lower than the AIC of the model without random intercepts. 

While I do not how how to *test* if including a random effect is justified, there are often situations, which require to test exactly which random effect structure is best. When doing this, it is important to use *restricted
maximum likelihood* (`REML = TRUE` or `method = REML`) rather than maximum likelihood [see @pinheiro2000mixedmodels; @winter2019statistics, 226].

```{r lmm9b, message=F, warning=F}
# generate models with 2 different random effect structures
ma.lmer = lmer(Prepositions ~ Date + (1|Genre), REML = T, data = lmmdata)
mb.lmer = lmer(Prepositions ~ Date + (1 + Date | Genre), REML = T, data = lmmdata)
# compare models
anova(ma.lmer, mb.lmer, test = "Chisq", refit = F)
```

The model comparison shows that the model with the more complex random effect structure has a significantly better fit to the data compared with the model with the simpler random effect structure. However, we will continue with the model with the simpler structure because this is just an example and we want to keep it simple here. 


<div class="warning" style='padding:0.1em; background-color:#f2f2f2; color:#51247a'>
<span>
<p style='margin-top:1em; text-align:center'>
<b>NOTE</b><br><br>In a real analysis, we would switch to a model with random intercepts and random slopes for Genre because it has a significantly better fit to the data.</p>
<p style='margin-left:1em;'>
</p></span>
</div>

<br>


## Model Fitting

After having determined that including a random effect structure is justified, we can continue by fitting the model and including diagnostics as we go. Including diagnostics in the model fitting process can save time and prevent relying on models which only turn out to be unstable if we would perform the diagnostics after the fact.

We begin fitting our model by adding *Date* as a fixed effect and compare this model to our mixed-effects base-line model to see if *Date* improved the model fit by explaining variance and if *Date* significantly correlates with our dependent variable (this means that the difference between the models is the effect (size) of *Date*!)


```{r lmm10}
m1.lmer <- lmer(Prepositions ~ (1|Genre) + Date, REML = T, data = lmmdata)
anova(m1.lmer, m0.lmer, test = "Chi")
```

```{r lmm10b, echo = F}
mc1 <- anova(m1.lmer, m0.lmer, test = "Chi")
```


The model with *Date* is the better model (significant p-value and lower AIC). The significant p-value shows that *Date* correlates significantly with *Prepositions* ($\chi$^2^(`r as.vector(unlist(mc1))[14][1]`): `r mc1$Chisq[2]`, p  = `r round(as.vector(unlist(mc1))[16][1], 5)`) . The $\chi$^2^ value here is labeled *Chisq* and the degrees of freedom are calculated by subtracting the smaller number of DFs from the larger number of DFs.

We now test whether *Region* should also be part of the final minimal adequate model. The easiest way to add predictors is by using the `update` function (it saves time and typing).

```{r lmm11}
# generate model
m2.lmer <- update(m1.lmer, .~.+ Region)
# test vifs
car::vif(m2.lmer)
# compare models                
anova(m2.lmer, m1.lmer, test = "Chi")
```

Three things tell us that *Region* should not be included: 

1. the AIC does not decrease, 

2. the BIC increases(!), and 

3. the p-value is higher than .05. 

This means, that we will continue fitting the model without having *Region* included. Well... not quite - just as a note on including variables: while *Region* is not significant as a main effect, it must still be included in a model if it were part of a significant interaction. To test if this is indeed the case, we fit another model with the interaction between *Date* and *Region* as predictor.

```{r lmm12}
# generate model
m3.lmer <- update(m1.lmer, .~.+ Region*Date)
# extract vifs
car::vif(m3.lmer)
# compare models                
anova(m3.lmer, m1.lmer, test = "Chi")
```

Again, the high p-value and the increase in AIC and BIC show that we have found our minimal adequate model with only contains *Date* as a main effect. In a next step, we can inspect the final minimal adequate model, i.e. the most parsimonious (the model that explains a maximum of variance with a minimum of predictors).

```{r lmm13}
# inspect results
summary(m1.lmer)
```


## Effect Sizes

We will now extract effect sizes (in the example: the effect size of *Date*) and calculate normalized effect size measures (this effect size measure works for all fixed effects). When you have factorial design, you can take the square root of the squared t-value divided by the t-value squared plus the degrees of freedom to calculate the effect size: 

\begin{equation}

r = \sqrt{ \frac{ t^2}{(t^2 + df) } } = \sqrt{ \frac{ 3.99^2}{(3.99^2 + 520) } } = `r round(sqrt(3.99^2/(3.99^2+520)), 3)`

\end{equation}

<br>

<div class="warning" style='padding:0.1em; background-color:#f2f2f2; color:#51247a'>
<span>
<p style='margin-top:1em; text-align:center'>
<b>NOTE</b><br><br>Two words of warning though: <br>br>1. In our case, the effect we are interested in is not factorial but continuous which means that we should not use this effect size measure. We only show this here as an example for how you can calculate the effect size measure r.<br><br>2. Only apply this function to main effects that are not involved in interactions as they are meaningless because the amount of variance explained by main effects involved in interactions is unclear [@field2012discovering 641].</p>
<p style='margin-left:1em;'>
</p></span>
</div>


<br>

```{r lmm21a, echo = F, eval = F, message=F, warning=F}
# WARNING: disfunct!
# calculate effect size 
#ggeffects::ggpredict(m5.lme, terms = "Date")
effects::predictorEffects(m1.lmer)
```

```{r lmm21g}
sjPlot::tab_model(m1.lmer)
```

***

<br>

<div class="warning" style='padding:0.1em; background-color:#f2f2f2; color:#51247a'>
<span>
<p style='margin-top:1em; text-align:center'>
<b>NOTE</b><br><br>The R^2^ values of the summary table are incorrect (as indicated by the missing conditional R^2^ value). The more appropriate conditional and marginal coefficient of determination for generalized mixed-effect models can be extracted using the `r.squaredGLMM` function from the `MuMIn` package [@MuMIn].</p>
<p style='margin-left:1em;'>
</p></span>
</div>


<br>

The *marginal R^2^* (marginal coefficient of determination) represents the variance explained by the fixed effects while the *conditional R^2^* is interpreted as a variance explained by the entire model, including both fixed and random effects [@barton2020mumin].

The respective call for the model is:


```{r lmm21c, warning = F, message=F}
# extract R2s
r.squaredGLMM(m1.lmer)
```

The effects can be visualized using the `plot_model` function from the `sjPlot` package [@sjPlot].

```{r lmm21d}
sjPlot::plot_model(m1.lmer, type = "pred", terms = c("Date")) +
  # show uncentered date rather than centered date
  scale_x_continuous(name = "Date", 
                     breaks = seq(-500, 300, 100), 
                     labels = seq(1150, 1950, 100))
```

While we have already shown that the effect of *Date* is significant, it is small which means that the number of prepositions per text does not correlate very strongly with time. This suggests that other factors that are not included in the model also impact the frequency of prepositions (and probably more meaningfully, too).


## Model Diagnostics

We now create diagnostic plots. What we wish to see in the diagnostic plots is a cloud of dots in the middle of the window without any structure. What we do not want to see is a funnel-shaped cloud because this indicates an increase of the errors/residuals with an increase of the predictor(s) (because this would indicate heteroscedasticity) [@pinheiro2000mixedmodels 182].

```{r lmm22}
# start plotting
par(mfrow = c(2, 2))           # display plots in 2 rows and 2 columns
plot(m1.lmer, pch = 20, col = "black", lty = "dotted"); par(mfrow = c(1, 1))
```

What a wonderful unstructured cloud - the lack of structure tells us that the model is "healthy" and does not suffer from heteroscedasticity. 

```{r lmm26}
# observed responses versus the within-group fitted values
plot(m1.lmer, Prepositions ~ fitted(.), id = 0.05, adj = -0.3, 
     xlim = c(80, 220), cex = .8, pch = 20, col = "blue")
```

Although some data points are named, the plot does not show any structure, like a funnel, which would have been problematic. 

## Reporting Results 

Before we do the write-up, we have a look at the model summary as this will provide us with at least some of the parameters that we want to report. 


```{r lmm28}
sjPlot::tab_model(m1.lmer)
```

***

<br>

<div class="warning" style='padding:0.1em; background-color:#f2f2f2; color:#51247a'>
<span>
<p style='margin-top:1em; text-align:center'>
<b>NOTE</b><br><br>The R^2^ values of the summary table are incorrect (as indicated by the missing conditional R^2^ value). The more appropriate conditional and marginal coefficient of determination for generalized mixed-effect models can be extracted using the `r.squaredGLMM` function from the `MuMIn` package [@MuMIn].</p>
<p style='margin-left:1em;'>
</p></span>
</div>

<br>


The respective call for the model is:

```{r lmm28b}
r.squaredGLMM(m1.lmer)
```


We can use the `reports` package [@report] to summarize the analysis.


```{r lme_report, message=F, warning = F}
report::report(m1.lmer)
```

We can use this output to write up a final report: 

A mixed-effect linear regression model which contained the genre of texts as random effect was fit to the data in a step-wise-step up procedure. Due to the presence of outliers in the data, weights were included into the model which led to a significantly improved model fit compared to an un-weight model ($\chi$^2^(2): 39.17, p: 0.0006). The final minimal adequate model performed significantly better than an intercept-only base-line model ($\chi$^2^(1): 12.44, p =.0004) and showed that the frequency of prepositions increases significantly but only marginally with the date of composition (Estimate: 0.02, CI: 0.01-0.03, p < .001, marginal R^2^ =  0.0174, conditional R^2^ =  0.4324). Neither the region where the text was composed nor a higher order interaction between genre and region significantly correlated with the use of prepositions in the data. 



[Back to top](#introduction)



