# (Different) Types of Regression


This week, we continue with regression analyses and expand on last week by using a regression to analyse a nominal dependent variable. Regression models that are fit to nominal dependent variables are called *(binomial) logistic regressions*.

**Preparation and session set up**

In this tutorial, we will need the following packages; please install them by running the code below if you have not installed them already. 

```{r prep0, eval = F, warning=F, message=F}
# install
install.packages("car")
install.packages("dplyr")
install.packages("car")
install.packages("flextable")
install.packages("gridExtra")
install.packages("ggplot2")
install.packages("rms")
install.packages("sjPlot")
```

Once you have installed the packages (either now or before), we activate them as shown below.

```{r prep1, echo=T, eval=T, warning=F, message=F}
# set options
options("scipen" = 100, "digits" = 12) # suppress math annotation
# load packages
library(dplyr)
library(car)
library(flextable)
library(gridExtra)
library(ggplot2)
library(rms)
library(sjPlot)
```


Once you have initiated the session by executing the code shown above, you are good to go.


## Basic Concepts

Logistic regression is a multivariate analysis technique that builds on linear regression but takes nominal dependent variables (rather than a numeric dependent variable) [@harrell2015regression]. The difference requires that the linear regression must be modified in certain ways to avoid predicting  outcomes that do not make sense. For example, a linear model that is applied to a nominal dependent variable (e.g., 0 = not pregnant, 1 = pregnant) would predict negative values or values greater than 1 when there are only 2 possible outcomes. 

To avoid this, logistic regressions work on the probabilities of an outcome (the likelihood), rather than the outcome itself. In addition, the likelihoods on which the logistic regression works is logged (logarithmized) in order to avoid making nonsensical predictions. 

To understand what this mean, we will use a very simple example. In this example, we want to see whether the height of men affect their likelihood of being in a relationship. The data we use represents a data set consisting of two variables: height and relationship.

```{r blm1, echo=F, message=FALSE, warning=FALSE}
# set seed
set.seed(12345)
# generates 20 values, with mean of 30 & s.d.=2
bodyheight=rnorm(20,175,20) 
# sorts these values in ascending order
bodyheight=sort(round(bodyheight, 1)) 
# assign 'survival' to these 20 individuals non-randomly
# most mortality occurs at smaller body size
relationship=c(0,0,0,0,0,1,0,1,0,0,1,1,0,1,1,1,0,1,1,1) 
# saves data frame with two columns: body size, survival
blrdata=as.data.frame(cbind(bodyheight,relationship)) 
# inspect data
blrdata %>%
  as.data.frame() %>%
  flextable() %>%
  flextable::set_table_properties(width = .75, layout = "autofit") %>%
  flextable::theme_zebra() %>%
  flextable::fontsize(size = 12) %>%
  flextable::fontsize(size = 12, part = "header") %>%
  flextable::align_text_col(align = "center") %>%
  flextable::border_outer()
```

Let us now fit a linear regression model to the data and see what happens if we plot the predictions of the model.

```{r blm1b, echo=F, message=FALSE, warning=FALSE}
# generate models
lm1 <- lm(relationship ~ bodyheight, data = blrdata)
blr1 <- glm(relationship ~ bodyheight, family = binomial, data = blrdata)
blrdata <- blrdata %>%
  dplyr::mutate(Pred_lm = predict(lm1, blrdata),
                Pred_blm = predict(blr1, blrdata)) %>%
  dplyr::mutate(Prob_lm = plogis(predict(lm1, blrdata)),
                Prob_blm = plogis(predict(blr1, blrdata)),
                Pred_blm2 = ifelse(predict(blr1, blrdata, type = "response") >= .5, 1, 0))
# plot 
ggplot(blrdata, aes(x = bodyheight, y =  relationship)) +
  geom_point() +
  geom_abline(intercept = coef(lm1)[1], slope = coef(lm1)[2], color = "red") +
  geom_point(aes(x = bodyheight, y = Pred_lm), color = "blue") +
  theme_bw(base_size = 10)+
  coord_cartesian(ylim = c(-0.2, 1.2), xlim = c(125, 225)) +
  scale_y_continuous(breaks=seq(0, 1, 1), labels = c("Single", "Relationship")) +
  guides(fill = FALSE) +
  labs(title = "Predictions and regression line of lm.", x = "Height", y = "")
```




You can check this by logging the values from -5 to 5 using the `plogis` function as shown below.

```{r blm0, message=FALSE, warning=FALSE}
round(plogis(-5:5), 5)
```

If we visualize these logged values, we get values that are between 0 and 1 and follow an S-shaped curve which mirrors the underlying logistic function.

```{r blm0b, echo = F, message=FALSE, warning=FALSE}
plogis(-5:5) %>%
  as.data.frame() %>%
  dplyr::mutate(id = 1:n()) %>%
  dplyr::rename(values = 1) %>%
  ggplot(aes(y = values, x = id)) +
  geom_line(color = "gray50", size = 1) +
  geom_point(color = "red", size = 2)+
  labs(x = "", y = "") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```




Let us now fit a binomial logistic regression model to the data and see what happens if we plot the predictions of the models as well as the probabilities provided by the logistic model.

```{r blm1f, echo=F, message=FALSE, warning=FALSE}
# generate models
lm1 <- lm(relationship ~ bodyheight, data = blrdata)
blr1 <- glm(relationship ~ bodyheight, family = binomial, data = blrdata)
blrdata <- blrdata %>%
  dplyr::mutate(Pred_lm = predict(lm1, blrdata),
                Pred_blm = predict(blr1, blrdata)) %>%
  dplyr::mutate(Prob_lm = plogis(predict(lm1, blrdata)),
                Prob_blm = plogis(predict(blr1, blrdata)),
                Pred_blm2 = ifelse(predict(blr1, blrdata, type = "response") >= .5, 1, 0))
# plot 
p2 <- ggplot(blrdata, aes(x = bodyheight, y =  relationship)) +
  geom_point() +
  geom_abline(intercept = coef(blr1)[1], slope = coef(blr1)[2], color = "red", size = .5) +
  geom_point(aes(x = bodyheight, y = Pred_blm), color = "blue") +
  theme_bw(base_size = 10)+
  coord_cartesian(ylim = c(-0.2, 1.2), xlim = c(125, 225)) +
  scale_y_continuous(breaks=seq(0, 1, 1), labels = c("Single", "Relationship")) +
  guides(fill = FALSE) +
  labs(title = "Predictions and regression line of blm.", x = "Height", y = "")


p3 <- ggplot(blrdata, aes(x = bodyheight, y =  relationship)) +
  geom_point() +
  geom_point(aes(x = bodyheight, y = Prob_blm), color = "blue") +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), 
              se = FALSE, color = "red", size = .5) +
  geom_segment(aes(xend = bodyheight, yend = Prob_blm), color = "red", alpha = .2) +
  theme_bw(base_size = 10) +
  coord_cartesian(ylim = c(-0.2, 1.2), xlim = c(125, 225)) +
  scale_y_continuous(breaks=seq(0, 1, 1), labels = c("Single", "Relationship")) +
  guides(fill = FALSE) +
  labs(title = "Logged Probabilities and logged regression line of blm.", x = "Height", y = "")

# show plot
ggpubr::ggarrange(p2, p3, nrow = 2)
```

The left panel of the figure above shows that a linear model would predict values for the relationship status, which represents a factor (0 = Single and 1 = In a Relationship), that are nonsensical because values above 1 or below 0 do not make sense. 

The center panel shows the predictions of a logistic regression and we see that a logistic regression also has an intercept and a (very steep) slope but that the regression line also predicts values that are above 1 and below 0. 

However, when we log the predicted values we these predicted values are transformed into probabilities with values between 0 and 1. And the logged regression line has a S-shape which reflects the logistic function. Furthermore, we can then find the optimal line (the line with the lowest residual deviance) by comparing the sum of residuals - just as we did for a simple linear model and that way, we find the regression line for a logistic regression. 

Thus, in contrast to a linear regression, which predicts actual values, such as the frequencies of prepositions in a certain text, a logistic regression predicts *probabilities* of events (for example, being in a relationship) rather than actual values. 


## Example: EH in Kiwi English

To exemplify how to implement a logistic regression in R [see @agresti1996introduction; @agresti2011categorical for very good and thorough introductions to this topic], we will analyze the use of the discourse particle *eh* in New Zealand English and test which factors correlate with its occurrence. The data set represents speech units in a corpus that were coded for the speaker who uttered a given speech unit, the gender, ethnicity, and age of that speaker and whether or not the speech unit contained an *eh*. To begin with, we clean the current work space, set option, install and activate relevant packages, load customized functions, and load the example data set.

```{r blm3}
# load data
blrdata  <- base::readRDS(url("https://slcladal.github.io/data/bld.rda", "rb"))
```

```{r blm4, echo = F}
# inspect data
blrdata %>%
  as.data.frame() %>%
  head(15) %>%
  flextable() %>%
  flextable::set_table_properties(width = .75, layout = "autofit") %>%
  flextable::theme_zebra() %>%
  flextable::fontsize(size = 12) %>%
  flextable::fontsize(size = 12, part = "header") %>%
  flextable::align_text_col(align = "center") %>%
  flextable::set_caption(caption = "First 15 rows of the blrdata.")  %>%
  flextable::border_outer()
```

The summary of the data show that the data set contains 25,821 observations of five variables. The variable `ID` contains strings that represent a combination file and speaker of a speech unit. The second variable represents the gender, the third the age, and the fourth the ethnicity of speakers. The fifth variable represents whether or not a speech unit contained the discourse particle *eh*. 

Next, we factorize the variables in our data set. In other words, we specify that the strings represent variable levels and define new reference levels because as a default `R` will use the variable level which first occurs in alphabet ordering as the reference level for each variable, we redefine the variable levels for Age and Ethnicity.

```{r blm5}
blrdata <- blrdata %>%
  # factorize variables
  dplyr::mutate_if(is.character, factor) %>%
  # relevel Age (Reference = Young) and Ethnicity (Reference= Pakeha))
  dplyr::mutate(Age = relevel(Age, "Young"),
                Ethnicity = relevel(Ethnicity, "Pakeha"))
```

After preparing the data, we will now plot the data to get an overview of potential relationships between variables.

```{r blm6, message=FALSE, warning=FALSE}
blrdata %>%
  dplyr::mutate(EH = ifelse(EH == "0", 0, 1)) %>%
  ggplot(aes(Age, EH, color = Gender)) +
  facet_wrap(~Ethnicity) +
  stat_summary(fun = mean, geom = "point") +
  stat_summary(fun.data = mean_cl_boot, geom = "errorbar", width = 0.2) +
  theme_bw(base_size = 10) +
  theme(legend.position = "top") +
  labs(x = "", y = "Observed Probabilty of eh") +
  scale_color_manual(values = c("gray20", "gray70"))
```

With respect to main effects, the Figure above indicates that men use *eh* more frequently than women, that young speakers use it more frequently compared with old speakers, and that speakers that are descendants of European settlers (Pakeha) use *eh* more frequently compared with Maori (the native inhabitants of New Zealand).

The plots in the lower panels do not indicate significant interactions between use of *eh* and the Age, Gender, and Ethnicity of speakers. In a next step, we will start building the logistic regression model.

## Model Building

As a first step, we need to define contrasts and use the `datadist` function to store aspects of our variables that can be accessed later when plotting and summarizing the model. Contrasts define what and how variable levels should be compared and therefore influences how the results of the regression analysis are presented. In this case, we use treatment contrasts which are in-built. Treatment contrasts mean that we assess the significance of levels of a predictor against a baseline which is the reference level of a predictor. @field2012discovering [414-427] and @gries2021statistics provide very good and accessible explanations of contrasts and how to manually define contrasts if you would like to know more. 

```{r blm7, message=F, warning=F}
# set contrasts
options(contrasts  =c("contr.treatment", "contr.poly"))
# extract distribution summaries for all potential variables
blrdata.dist <- datadist(blrdata)
# store distribution summaries for all potential variables
options(datadist = "blrdata.dist")
```

Next, we generate a minimal model that predicts the use of *eh* solely based on the intercept.

```{r blm8}
# baseline glm model
m0.glm = glm(EH ~ 1, family = binomial, data = blrdata)
```

## Model Fitting

We will now start with the model fitting procedure. In the present case, we will use a manual step-wise step-up procedure during which predictors are added to the model if they significantly improve the model fit. In addition, we will perform diagnostics as we fit the model at each step of the model fitting process rather than after the fitting.

We will test two things in particular: whether the data has incomplete information or complete separation, if the model suffers from (multi-)collinearity, and if the predictor is significant. 

**Incomplete information**  means that the data does not contain all combinations of the predictor or the dependent variable. 

**Complete separation** is important because if the data does not contain cases of all combinations, the model will assume that it has found a perfect predictor. In such cases the model overestimates the effect of that that predictor and the results of that model are no longer reliable. For example, if *eh* was only used by young speakers in the data, the model would jump on that fact and say *Ha! If there is an old speaker, that means that that speaker will never ever and under no circumstances say *eh* - I can therefore ignore all other factors!* 

**Multicollinearity** means that predictors correlate and have shared variance. This means that whichever predictor is included first will take all the variance that it can explain and the remaining part of the variable that is shared will not be attributed to the other predictor. This may lead to reporting that a factor is not significant because all of the variance it can explain is already accounted for. However, if the other predictor were included first, then the original predictor would be returned as insignificant. This means that- depending on the order in which predictors are added - the results of the regression can differ dramatically and the model is therefore not reliable. Multicollinearity is actually a very common problem and there are various ways to deal with it but it cannot be ignored (at least in regression analyses).

**Significance**: if a model with a predictor is not better (has no better fit) compared to a model without that predictor, then we do not include the predictor in the model. In other words, we continue fitting the model without that predictor. However, If a variable is insignificant but it is part of a significant interaction, we also need to include the insignificant variable. 

**Step-wise model fitting**

Adding predictors one-by-one is called step-wise model fitting. It is called step-wise step up if we add predictors one-by-one and step-wise step-down if we remove predictors one-by-one.

We will start by adding *Age* to the minimal adequate model. 

```{r blm9}
# check incomplete information
ftable(blrdata$Age, blrdata$EH)
# add age to the model
m1.glm = glm(EH ~ Age, family = binomial, data = blrdata)
# check if adding Age significantly improves model fit
anova(m1.glm, m0.glm, test = "Chi")
```

As the data does not contain incomplete information, the vif values are below 3, and adding *Age* has significantly improved the model fit (the p-value of the ANOVA is lower than .05). We therefore proceed with *Age* included.

We continue by adding *Gender*. We add a second ANOVA test to see if including Gender affects the significance of other predictors in the model. If this were the case - if adding Gender would cause Age to become insignificant - then we could change the ordering in which we include predictors into our model.

```{r blm10}
ftable(blrdata$Gender, blrdata$EH)
m2.glm <- update(m1.glm, . ~ . +Gender)
vif(m2.glm)
anova(m1.glm, m2.glm, test = "Chi")
Anova(m2.glm, test = "LR")
```


Again, including *Gender* significantly improves model fit and the data does not contain incomplete information or complete separation. Also, including *Gender* does not affect the significance of *Age*. Now, we include *Ethnicity*.


```{r blm11}
ftable(blrdata$Ethnicity, blrdata$EH)
m3.glm <- update(m2.glm, . ~ . +Ethnicity)
vif(m3.glm)
anova(m2.glm, m3.glm, test = "Chi")
```

Since adding *Ethnicity* does not significantly improve the model fit, we do not need to test if its inclusion affects the significance of other predictors. We continue without *Ethnicity* and include the interaction between *Age* and *Gender*.

```{r blm12}
ftable(blrdata$Age, blrdata$Gender, blrdata$EH)
m4.glm <- update(m2.glm, . ~ . +Age*Gender)
vif(m4.glm)
anova(m2.glm, m4.glm, test = "Chi")
```

The interaction between *Age* and *Gender* is not significant which means that men and women do not behave differently with respect to their use of `EH` as they age. Also, the data does not contain incomplete information and the model does not suffer from multicollinearity - the predictors are not collinear. We can now include if there is a significant interaction between *Age* and *Ethnicity*.

```{r blm13}
ftable(blrdata$Age, blrdata$Ethnicity, blrdata$EH)
m5.glm <- update(m2.glm, . ~ . +Age*Ethnicity)
vif(m5.glm)
anova(m2.glm, m5.glm, test = "Chi")
```

Again, no incomplete information or multicollinearity and no significant interaction. Now, we test if there exists a significant interaction between *Gender* and *Ethnicity*.

```{r blm14}
ftable(blrdata$Gender, blrdata$Ethnicity, blrdata$EH)
m6.glm <- update(m2.glm, . ~ . +Gender*Ethnicity)
vif(m6.glm)
anova(m2.glm, m6.glm, test = "Chi")
```


As the interaction between *Gender* and *Ethnicity* is not significant, we continue without it. In a final step, we include the three-way interaction between *Age*, *Gender*, and *Ethnicity*.

```{r blm15}
ftable(blrdata$Age, blrdata$Gender, blrdata$Ethnicity, blrdata$EH)
m7.glm <- update(m2.glm, . ~ . +Age*Gender*Ethnicity)
vif(m7.glm)
anova(m2.glm, m7.glm, test = "Chi")
```

We have found our final minimal adequate model because the 3-way interaction is also insignificant. As we have now arrived at the final minimal adequate model (m2.glm), we continue by checking the effect sizes of the predictors in our final model (Age and Gender).


## Effect Size

In a next step, we can use the `plot_model` function from the `sjPlot` package [@sjPlot] to visualize the effects.

```{r blm28, message=F, warning=F}
sjPlot::plot_model(m2.glm, 
                   type = "pred", 
                   terms = c("Age", "Gender"), 
                   axis.lim = c(0, 1)) +
  theme(legend.position = "top") +
  labs(x = "", y = "Predicted Probabilty of eh", title = "") +
  scale_color_manual(values = c("gray20", "gray70")) +
  theme_bw()
```


## Summarizing Results

As a final step, we summarize our findings in tabulated form.

```{r blm33, message=F, warning=F}
sjPlot::tab_model(m2.glm)
```



We can use the `reports` package [@report] to summarize the analysis.


```{r blr_report, message=F, warning = F}
report::report(m2.glm)
```

We can use this output to write up a final report: 

We fitted a logistic model (estimated using ML) to predict the use of the utterance-final discourse particle *eh* with Age and Gender (formula: EH ~ Age + Gender). The model's explanatory power is weak (Tjur's R^2^ = 0.03). The model's intercept, corresponding to Age = Young and Gender = Men, is at -0.23 (95% CI [-0.28, -0.19], p < .001). Within this model:

* The effect of Age [Old] is statistically significant and negative (beta = -0.83, 95% CI [-0.90, -0.77], p < .001; Std. beta = -0.83, 95% CI [-0.90, -0.77])

* The effect of Gender [Women] is statistically significant and negative (beta = -0.42, 95% CI [-0.47, -0.37], p < .001; Std. beta = -0.42, 95% CI [-0.47, -0.37])

Standardized parameters were obtained by fitting the model on a standardized version of the dataset. 95% Confidence Intervals (CIs) and p-values were computed using 




[Back to top](#introduction)



