---
title: "Untitled"
author: "Martin Schweinberger"
date: '2022-08-20'
output: html_document
---


## Preparation and session set up{-}

This tutorial is based on R. If you have not installed R or are new to it, you will find an introduction to and more information how to use R [here](https://slcladal.github.io/intror.html). For this tutorials, we need to install certain *packages* from an R *library* so that the scripts shown below are executed without errors. Before turning to the code below, please install the packages by running the code below this paragraph. If you have already installed the packages mentioned below, then you can skip ahead and ignore this section. To install the necessary packages, simply run the following code - it may take some time (between 1 and 5 minutes to install all of the libraries so you do not need to worry if it takes some time).

```{r prep1, echo=T, eval = F, message=FALSE, warning=FALSE}
# install packages
install.packages("dplyr") 
install.packages("ggplot2") 
install.packages("tidyr") 
install.packages("flextable") 
install.packages("e1071") 
install.packages("lawstat") 
install.packages("fGarch") 
install.packages("gridExtra") 
install.packages("cfa") 
# install klippy for copy-to-clipboard button in code chunks
install.packages("remotes")
remotes::install_github("rlesur/klippy")
```

Now, we load the packages that we will need in this tutorial.

```{r qq01, message=F, warning=F}
# load packages
library(dplyr)        # for data processing
library(ggplot2)      # for data vis
library(tidyr)        # for data transformation
library(flextable)    # for creating tables
library(e1071)        # for checking assumptions
library(lawstat)      # for statistical tests
library(fGarch)       # for statistical tests
library(gridExtra)    # for plotting
library(cfa)          # for stats
# activate klippy for copy-to-clipboard button
klippy::klippy()
```